---
title: "R functions to Galaxy tools"
author: "Lodewic van Twillert"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r libraries}
# install.packages(RGalaxy)

# library(RGalaxify) 
# Alternatively, source all code
for (fname in dir("../R/")) source(file.path("../R", fname))
library(XML)
library(tools)
# library(RGalaxy) # We still need some core functions that we haven't edited yet
```

```{r tool generation functions}
addCommandXML <- function(funcName, funcInfo, xml) {
  # Make command text
  commandText <- paste(funcName, ".R\n", sep="")
      
  for (nm in names(funcInfo))
  {
      commandText <- paste(commandText, "       ",
          sprintf("#if str($%s).strip() != \"\":\n", nm),
          "          ", sprintf("--%s=\"$%s\"", nm, nm),
          "\n       #end if\n",
          sep="")
  }
  
  command.pre <- paste0(c("\n    ######### one important path:",
                          "    ######### 1. path to tool installation directory",
                          "    export TOOL_INSTALL_DIR='${__tool_directory__}' &&\n"),
                          collapse = "\n")
      
  commandText <- gsub(funcName, paste0("'$__tool_directory__/", funcName, "'"), commandText)
  commandText <- paste0(c(command.pre, "\n    Rscript ", commandText), collapse = "")
  
  commandNode <- newXMLNode("command",
                            parent = xml, # genest in tool
                            attrs = list(detect_errors = "exit_code"),
                            newXMLCDataNode(commandText))
  return(xml)
}

addParamsXML <- function(funcName, funcInfo, xml) {
  functionToGalaxify <- get(funcName)
  
  inputsNode <- newXMLNode("inputs", parent=xml)
  outputsNode <- newXMLNode("outputs", parent=xml)

  # Loop over Galaxy Parameter Class arguments of the given function
  #   Each parameter is classified with an if/else chain and
  #   the resulting xml-tags are defined accordingly
  for (nm in names(funcInfo))
{
    item <- funcInfo[nm][[1]]
    galaxyItem <- eval(formals(functionToGalaxify)[nm][[1]])
    if (!item$type == "GalaxyOutput")
    {
        paramNode <- newXMLNode("param", parent=inputsNode)
        if (galaxyItem@required)
        {
            validatorNode <- newXMLNode("validator", parent=paramNode)
            xmlAttrs(validatorNode)["type"] <- "empty_field"
            xmlAttrs(validatorNode)["message"] <- galaxyItem@requiredMsg
            xmlAttrs(paramNode)['optional'] <- 'false'
        } else {
            validatorNode <- newXMLNode("validator", parent=paramNode)
            xmlAttrs(validatorNode)["type"] <- "empty_field"
            ##dummyParam <- GalaxyParam()
            xmlAttrs(validatorNode)["message"] <-
                galaxyItem@requiredMsg
            xmlAttrs(paramNode)['optional'] <- 'true'
        }
        if (item$type == "GalaxyInputFile")
        {
            xmlAttrs(paramNode)["optional"] <-
                tolower(
                    as.character(!eval(formals(functionToGalaxify)[[nm]])@required))
        }
        
        xmlAttrs(paramNode)["name"] <- nm
        type <- RtoGalaxyTypeMap[[item$type]]
        if (item$type == "GalaxyInputFile")
        {
            type <- "data"
            if (length(galaxyItem@formatFilter))
                xmlAttrs(paramNode)["format"] <- galaxyItem@formatFilter
            else
                xmlAttrs(paramNode)["format"] <- "data" #default format of an input file
        }
        if (item$length > 1 || "GalaxySelectParam" %in% class(galaxyItem))
        {
            type <- "select"
        }
        xmlAttrs(paramNode)["type"] <- type

        if(!is.null(item$default))
            xmlAttrs(paramNode)["value"] <- eval(item$default)
        else
            if (type %in% c("integer", "float"))
                xmlAttrs(paramNode)["value"] <- ""

        xmlAttrs(paramNode)["help"] <- getHelpFromText(rd, nm)
        
        if (length(galaxyItem@label)) ## this really should always be true!
            item$label <- galaxyItem@label

        if ( galaxyItem@required){
            item$label <- paste("[required]", item$label)
        }
        

        if (type == "boolean")
        {
            if (length(galaxyItem@checked))
                xmlAttrs(paramNode)['checked'] <-
                    tolower(as.character(galaxyItem@checked))
        }

        if (type == "text") {
            galaxyItem
            if (length(galaxyItem@size))
            xmlAttrs(paramNode)['size'] = as.character(galaxyItem@size)
        }

        if(type %in% c("integer", "float"))
        {
            if(length(galaxyItem@min))
                xmlAttrs(paramNode)['min'] <- as.character(galaxyItem@min)
            if(length(galaxyItem@max))
                xmlAttrs(paramNode)['max'] <- as.character(galaxyItem@max)
        }

        
        xmlAttrs(paramNode)['label'] <- item$label
        
        
        if (type=="select")
        {
            xmlAttrs(paramNode)['force_select'] <-
                as.character(galaxyItem@force_select)
            if (length(galaxyItem@display))
                xmlAttrs(paramNode)['display'] <-
                    as.character(galaxyItem@display)


            if (!is.null(item$selectoptions))
            {
                selectoptions <- eval(item$selectoptions)
                idx <- 1
                for (value in selectoptions)
                {
                    option <- names(selectoptions)[[idx]]
                    if (is.null(option)) option <- value
                    optionNode <- newXMLNode("option", option,
                        parent=paramNode)
                    xmlAttrs(optionNode)['value'] <- value
                    idx <- idx + 1
                }
                
            }

        }
        invisible(NULL)
        
    } else
    {
        dataNode <- newXMLNode("data", parent=outputsNode)
        if (is.null(item$default))
            gstop(sprintf("GalaxyOutput '%s' must have a parameter.", nm))
        galaxyOutput <- eval(item$default)
        xmlAttrs(dataNode)["format"] <- galaxyOutput@format
        xmlAttrs(dataNode)["name"] <- nm
        xmlAttrs(dataNode)["label"] <- as.character(galaxyOutput)
        
    }
  
  }
  return(xml)
}

addTestsXML <- function(funcName, funcInfo, xml,
                        testDataDir = "mytool/tools/test-data") {
  #testFileDest <- file.path(funcName)
  testsNode <- newXMLNode("tests", parent = xml)
  testNode <- newXMLNode("test", parent=testsNode)
  for (info in funcInfo)
  {
      cat(sprintf("Func info for %s: \n", info$param))
      print(funcInfo)
      if (info$type == "GalaxyOutput")
          testParamNode <- newXMLNode("output", parent=testNode)
      else
          testParamNode <- newXMLNode("param", parent=testNode)
      xmlAttrs(testParamNode)["name"] <- info$param
      if (length(info$type) > 0 && 
          info$type == "GalaxyOutput")
      {   
          # # Copy output file to the test-data directory in the same folder
          # # as the .xml tool definition
          # if (!exists("functionalTestDirectory")) {
          #   srcFile <- system.file("functionalTests", funcName, eval(info$default)[1],
          #                          package=package)
          # } else srcFile <- file.path(functionalTestDirectory, funcName, eval(info$default)[1])
          # if (!file.exists(file.path(testDataDir, eval(info$default)[1])))
          #     file.copy(srcFile, testDataDir)
          xmlAttrs(testParamNode)["file"] <- eval(info$default)[1]
      }
      # If GalaxyInputFile we now still expect testValues!
      if (info$type=="GalaxyInputFile") {
        if (is.null(info$testValues)) {
          gwarning("Setting input file for `%s` without extension as test data", info$param)
          xmlAttrs(testParamNode)['value'] <- info$param
        } else {
          ## for now, just assume one value
          xmlAttrs(testParamNode)['value'] <- info$testValues
        }
        # Copy input file to the test-data directory
        # if (!exists("functionalTestDirectory")) {
        #   srcFile <- system.file("functionalTests", funcName, xmlAttrs(testParamNode)['value'],
        #                          package=package)
        # } else srcFile <- file.path(functionalTestDirectory, funcName, xmlAttrs(testParamNode)['value'])
        # file.copy(srcFile, testDataDir)
      # If parameter is Not a GalaxyInputFile
      #   Note: GalaxyOutput never has a 'testValues' field.
      } else {
        if (!is.null(info$testValues) && length(info$testValues) > 0)
        {
          ## for now, just assume one value
          xmlAttrs(testParamNode)['value'] <- info$testValues
        }
      }
      
      
  }
  return(xml)
}

```

# Generating a Galaxy tool from local R scripts

This vignette is the first attempt to generate a Galaxy tool from an arbitrary
R script. The script is expected to include functions with roxygen-style 
documentation.

Also, these R functions should be lightly edited so that their expected 
parameters can be parsed by the functions from the [RGalaxy](https://bioconductor.org/packages/release/bioc/html/RGalaxy.html)
package.

The biggest shortcoming of the RGalaxy package in my opinion is that the generated
tool script relies on the used R function to be in a publically available package.
The Galaxy tools will not include the right requirements tags needed to publish 
the generated Galaxy tools to any Galaxy instance.

Our approach will use a lot of the functionality of the RGalaxy package, but 
rewritten in a way that we use arbitrary custom R functions. To see an example
workflow of how to generate tools using RGalaxy, see 
[this RGalaxy vignette](https://bioconductor.org/packages/release/bioc/vignettes/RGalaxy/inst/doc/RGalaxy-vignette.html). However, the vignette is not very easy to follow but
it shows some examples and descriptions that we won't go over in this document.

# Creating a basic galaxy tool

Your R scripts need to be self-contained function definitions, for now.
Later this should be extended to include tools that depend on other local functions
that may be loaded with a simple `source()` command.

The source R scripts can be given as a character vector of filenames, relative
to the current working directory.

## AddTwoNumbersWithTest

In this case we use the `addTwoNumbersWithTest.R`. This file includes one very
simple function defintion. All this script included is a function to sum two numbers and write the output to
a file. In Galaxy we always expect the output to be a file, or collection of files.
You can output R variables in .rds format, which may force you to change some
other functions to take a file input rather than just an R variable.

Notice the following about the function definition below,

  * [roxygen2](http://kbroman.org/pkg_primer/pages/docs.html)-style comments for function documentation
  * The input parameters are [Galaxy Parameter Classes](https://rdrr.io/bioc/RGalaxy/man/GalaxyClasses.html)
  * The output is written to a .txt file
  
The documentation is used to create a help section to display in Galaxy.

The input parameters are defined by Galaxy Parameter Classes, following the 
methods in the RGalaxy package. This allows galaxy to understand the expected
parameters and show them nicely in the Galaxy UI.

```{r}
# The addTwoNumbersTest.R file

#' Add to numbers
#' 
#' A test function to use as an example to create a galaxy tool.
#' It takes two numeric values and returns the sum to a .txt file.
#'
#' @param number1 First numeric value to sum
#' @param number2 Second numeric value to sum
#' @param sum Output file
#' 
#' @details
#' Both arguments are required and contain test values to
#' automatically generate tool tests with.
#' These tests are included in the output .xml tool definition, and
#' can be tested using Planemo's test command.
#'
#' @return The sum of two values
#' @export
#'
#' @examples
# outfile <- tempfile(pattern = "sum")
# addTwoNumbersWithTest(5, 5, sum = outfile)
# readLines(outfile)
#' 
addTwoNumbersWithTest <- 
  function(
    number1=GalaxyNumericParam(required=TRUE, testValues=5L),
    number2=GalaxyNumericParam(required=TRUE, testValues=5L),
    sum=GalaxyOutput("sum", "txt"))
  {
    cat(paste0(number1 + number2, "\n"), file=sum)
  }
```

You can still use this function straight in R, which is handy for testing but
in many cases you could just make it a habit of writing function definitions like
this right away!

```{r}
# Example of function output using normal R variables
outfile <- tempfile()
addTwoNumbersWithTest(number1 = 5, number2 = 5, sum = outfile)
readLines(outfile)
```

## Setting up tool output

The output will be written to a tool directory, `toolDir`.
You may have only 1 tool directory but still have different groups of tools, 
set by `toolGroup`. 

```{r Set parameters}
toolScripts <- c("addTwoNumbersWithTest.R")
toolDir <- "tools"
toolGroup <- "mytool"
toolPath <- file.path(toolDir, toolGroup)

# Create tool directory if it does not yet exist
if (!dir.exists(toolPath)) {
  warning(sprintf("Tool directory '%s' does not exist and will be created",
                  toolPath))
  dir.create(toolPath, recursive = T)
}
```

## Creating temporary package

Since our functions to transform into galaxy tools are not part of a package 
we can't just use `roxygen2::roxygenise()`. For us to create documentation the 
scripts need to be part of a package. So we create a temporary package for our functions.

```{r}
require(devtools)

# Create temporary package directory
tempPackageDir <- file.path(tempdir(), "TempGalaxyToolPackage")
# Initialize R package file structure
if (dir.exists(tempPackageDir)) unlink(tempPackageDir, recursive = T)
devtools::create(path = tempPackageDir,
                 rstudio = F)

# Copy scripts to package R directory
file.copy(from = toolScripts, to = file.path(tempPackageDir, "R"))
# Document R functions in the R directory
devtools::document(tempPackageDir)

# File directory example
list.files(tempPackageDir, recursive = T)
```

For every **documented** function we can create a galaxy tool.
Every function gets its own .Rd file in the `man/` directory. Using this
we can list the functions to galaxify.

```{r List functions to galaxify}
# List all functions to galaxify
#   Based on the criteria that they need to be documented
functionDocs <- list.files(file.path(tempPackageDir, "man"))
# Only 1 function in this case, otherwise we should create a loop.
funcName <- sapply(functionDocs, gsub, pattern = ".Rd", replacement = "")
cat(funcName,
    sep = "\n")
```

## Prepare documentation and function definition

Firstly we need to gather the documentation for our new tool/function.
This can be found in the `man/` directory of our temporary package. 

Here we print the contents of the .Rd documentation file created when we ran the
`devtools::document()` function. These files get created by the `roxygen2`package,
look up the documentation if you wish to learn more about the options and 
formatting.

```{r Get documentation for addTwoNumbersWithTest}
rd <- tools::parse_Rd(file.path(tempPackageDir, "man", functionDocs))
print(rd)
```

```{r Get definition of addTwoNumbersWithTest}
fullToolDir <- toolPath
scriptFileName <-  toolScripts
source(scriptFileName)
functionToGalaxify <- get(funcName) # Get the function to galaxify
funcInfo <- list()
print(formals(functionToGalaxify))
```

## Get function parameters

```{r}
if  (  length(names(formals(functionToGalaxify)))   > length(formals(functionToGalaxify)) )
  gstop("All arguments to Galaxy must be named.")

for (param in names(formals(functionToGalaxify)))
  funcInfo[[param]] <- getFuncInfo(functionToGalaxify, param)

# # print(funcInfo)
# if (!isTestable(funcInfo, funcName, package, functionalTestDirectory)) 
#   gwarning("Not enough information to create a functional test.")
  
if (!suppressWarnings(any(lapply(funcInfo,
  function(x)x$type=="GalaxyOutput"))))
{
  gstop(paste("You must supply at least one GalaxyOutput",
      "object."))
}
```

# Further work

This is still a work in progress, below are the next steps to rewrite.
All this code is taken from the RGalaxy package, or specifically from my
edited Github mirror of it [found here](https://github.com/Lodewic/RGalaxy).

Look for the `R/` repository and you'll find the `R/galaxy.R` script that contains
most of the functions we need. Although many of the functions we DONT need, 
so we will work on rewriting the required functions.

```{r Create functional test}
if (!dir.exists(file.path(toolPath, "test-data"))) {
  message("Functional test data directory does not exist. /test-data will be created but you will need to add your test input data files to this folder.")
  dir.create(file.path(toolPath, "test-data"))
}

funcTestSuccess <- CreateFunctionalTest(funcName, 
                                        functionalTestDirectory = file.path(toolPath, "test-data"))
```

```{r Create xml text}
xmlFileName <- file.path(fullToolDir, paste(funcName, "xml", sep="."))
unlink(xmlFileName)

xml <- newXMLNode("tool",
                  attrs = list(id = funcName,
                               name = getFriendlyName(funcName)))
descNode <- newXMLNode("description",
                       newXMLTextNode(paste0(tools:::.Rd_get_metadata(rd, "description"), 
                                             collapse = " ")),
                       parent=xml)

reqNode <- newXMLNode("requirements",
                      parent = xml)
reqOptparse <- newXMLNode("requirement", 
                          attrs = list(type = "package", version = "1.4.4"),
                          newXMLTextNode("r-optparse"),
                          parent = reqNode)
reqOptparse <- newXMLNode("requirement", 
                          attrs = list(type = "package", version = "1.22.0"),
                          newXMLTextNode("bioconductor-rgalaxy"),
                          parent = reqNode) 

xml <- addCommandXML(funcName, funcInfo, xml)
xml <- addParamsXML(funcName, funcInfo, xml)

helpText <- generateHelpText(rd)
helpNode <- newXMLNode("help", newXMLTextNode(helpText), parent=xml)
testsNode <- addTestsXML(funcName, funcInfo, xml)
citationNode <- newXMLNode("citations", parent=xml)
saveXML(xml, file=xmlFileName)
```

```{r Create R script runnable from cmd}
# Use the generate_runnable.R script 
# to make any function runnable from the command line:)
scriptOut <- RewriteToolScript(toolScripts[1], toolPath = toolPath)
```
